\bf Estimating the number of clusters using cross-validation

### 2016-06-17 cvclust.tex
revise section 9

In the \citet{tavazoie1999systematic} analysis, those authors performed
$k$-means clustering with $k = 30$; they found 23 of the clusters to be
uninterpretable, and they found 7 clusters to be meaningful. To compare our
clusters with the Tavazoie et al.\ clusters, we prepared a confusion matrix
comparing our clusters with the 7 interpretable Tavazoie clusters in
Table~\ref{table:confusion}.  (Appendix~\ref{app:confusion} contains a
confusion matrix for all 30 Tavazoie clusters, including the non-interpretable
ones.) Entry $(i,j)$ of the confusion matrix gives the number of genes in
Tavazoie's Cluster~$i$ and our Cluster~$j$.

@@

In this paper, we proposed a novel approach to estimate the number of
clusters. The intuition behind our proposed methods is to transfer the
unsupervised learning problem into supervised learning problem via novel form
of cross validation. Such approach is quite different from previous methods which 
utilize the within/between cluster dispersion or stability criterion for selecting the 
optimal $k$. Our method utilizes the connection between different dimensions (columns)
of data through the uniqueness of each cluster center. We proved the self-consistency 
for our proposed Gabriel cross-validation method as well as its asymptotic property with Gaussian noise,
and showed the robustness of our method by simulation. Our method has very good performance
in our limited simulation settings and real data application, and clearly the superior
one when data has heterogeneous variance or is heavy-tailed.

Besides no strong modeling assumption is required, our proposed method is robust against data set 
with variance heterogeneity, unequal number of observations, non-Gaussian noise and high-dimension. 
Such robustness is important in practice because for any given data, it's hard to tell whether
its clusters have different number of observations, is the variance equal for each cluster,
or what underlying noise distribution it has. The weakness of our proposed method is 
that its theory assumes only week correlation between ``predictor'' columns and ``response'' 
columns. In practice, many data sets don't exhibit high correlations between columns, where
our proposed method can be safely applied. In case the high correlation does exist, 
the correlation-corrected version of Gabriel cross-validation method in section $4.3$ can be used if we can 
assume common covariance structure. Other procedure such as leave out most columns for clustering 
may also be used to reduce its effect. However, the theory for such procedure has not
been fully developed and it could be the future research topic.

Another situation where our proposed method cannot be directly used is when the clusters 
are non-convex. In such situation, $k$-means itself doesn't work well, for example two 
concentric circles share the same cluster center \citep{hastie2009elements}. However, it's 
possible that our proposed Gabriel cross-validation method can be used on the transformed data set. In the
concentric circles case where spectral clustering is appropriate, our proposed method can be applied 
on the eigenvector subspace of the graph Laplacian matrix inside the spectral clustering algorithm
to find the optimal $k$. This can also be the future research topic.

@@

\section{Clustering scree plot}

@@

\section{Review of consistency results for clustering}

Some proposed methods provide theoretical consistency result for choosing $k$.
Assume data is a mixture of $G$ $p$-dimension clusters with equal prior,
identically distributed with common covariance matrix and finite fourth
moments in each dimension, \citet{sugar2003finding} shows the Jump method will
pick the correct $k$ under the conditions that cluster centers are
sufficiently separated and an appropriate transformation is used.
Specifically, let $\Delta\sqrt{p}$ denotes the minimum Euclidean distance
between cluster centers, \citet{sugar2003finding} shows that the jump will be
maximized at $k=G$ given that $\Delta > 6$ and the existence of a positive $Y$
such that
\[
  \left(\frac{p\Delta^2W}{9G}\right)^{-Y}
    + \left(
        W \left[\frac{2^{2H^*(X)}}{K^2_{max}2\pi e}\right]
        - (\frac{\Delta}{6})^2(1-W)
      \right)^{-Y} < 2
\]
and $\left(\frac{p\Delta^2W}{9G}\right)^{-Y} <1/2$, where $H^*(X)$ is the
minimum entropy of the cluster membership over each dimension and $W =1-
\frac{6^4V_{\mathbf{X}}}{(\Delta^2-36)^2}$ with $V_{\mathbf{X}} = \var\left(
  1/p ||\mathbf{X}-\mu_j||^2_{\Gamma^{-1}} \mid \mathbf{X}
\hspace{0.05in}\text{in jth cluster}\right)$

\cite{tibshirani2005cluster} proved that the prediction strength method is
consistent under the setting that the observations in $p$-dimension are
uniformly distributed within one unit from their respective cluster centers,
with minimum distance between the $G$ cluster centers is four unit. Under such
well-separated setting, they show that 
\[ ps(G) = 1+ o_p(1), \hspace{0.5in} \underset{k \geq G+1}{\text{sup}} ps(k)\leq \frac{2}{3}+ o_p(1) \]
therefore $\hat{k}$ is consistent in estimating $G$.  When data does come from
mixture of Gaussian distribution, the model-based method of
\citet{fraley2002model} is consistent in estimating $k$ in low dimension.

Let $\Psi$ denotes any given base-clustering algorithm (e.g $k$-means), and
$\Psi(Z,k)$ the clustering obtained by applying $\Psi$ on data $Z$ with
parameter $k$. Also let $k_0$ be the optimal number of cluster and
$d\{\Psi(Z_1,k),\Psi(Z_2,k)\}$ be the measure of distance between clustering
$\Psi(Z_1,k)$ and $\Psi(Z_2,k)$ where $Z_1$ and $Z_2$ are two independent
samples from population. \citet{wang2010consistent} shows that their proposed
method will consistently select $k = k_0$, given such $k_0$ exists and
$d\{\Psi(Z_1,k),\Psi(Z_2,k)\}$ converges to zero at proper rate $r_k$.
However, $k_0$ is defined as the most stable number for $\Psi$ among all the
possible $k$, instead of the number of components in mixture model or
well-separated compact regions commonly seen in literature. Therefore, their
consistent result actually is convergent result, i.e. the selected $k$
converges to its limit $k_0$ given such limit exists.

@@

\section{Confusion matrix} \label{app:confusion}

\begin{table}
%\captionsetup{justification=centering}
\caption{\label{table:confusion-complete} Complete confusion matrix comparing $5$ clusters to $30$ clusters }
\begin{tabular}{lccccc|c}
  \hline
 & Cluster 1 & Cluster 2 & Cluster 3 & Cluster 4 & Cluster 5 & Total \\ 
  \hline
  Cluster 1 & 0 & 0 & 1 & 161 & 2 & 164 \\ 
  Cluster 2 & 1 & 0 & 0 & 0 & 185 & 186 \\ 
  Cluster 3 & 0 & 0 & 91 & 11 & 2 & 104 \\ 
  Cluster 4 & 0 & 102 & 2 & 66 & 0 & 170 \\ 
  Cluster 5 & 152 & 0 & 0 & 0 & 0 & 152 \\ 
  Cluster 6 & 1 & 2 & 12 & 85 & 4 & 104 \\ 
  Cluster 7 & 1 & 10 & 83 & 7 & 0 & 101 \\ 
  Cluster 8 & 3 & 145 & 0 & 0 & 0 & 148 \\ 
  Cluster 9 & 142 & 0 & 0 & 0 & 5 & 147 \\ 
  Cluster 10 & 12 & 27 & 24 & 20 & 6 & 89 \\ 
  Cluster 11 & 0 & 5 & 81 & 8 & 0 & 94 \\ 
  Cluster 12 & 3 & 68 & 5 & 3 & 1 & 80 \\ 
  Cluster 13 & 0 & 1 & 12 & 81 & 5 & 99 \\ 
  Cluster 14 & 0 & 1 & 29 & 6 & 38 & 74 \\ 
  Cluster 15 & 60 & 53 & 2 & 0 & 0 & 115 \\ 
  Cluster 16 & 7 & 0 & 0 & 9 & 83 & 99 \\ 
  Cluster 17 & 9 & 0 & 72 & 1 & 1 & 83 \\ 
  Cluster 18 & 3 & 16 & 10 & 67 & 5 & 101 \\ 
  Cluster 19 & 62 & 2 & 0 & 1 & 8 & 73 \\ 
  Cluster 20 & 11 & 0 & 6 & 2 & 65 & 84 \\ 
  Cluster 21 & 1 & 13 & 6 & 44 & 6 & 70 \\ 
  Cluster 22 & 33 & 15 & 28 & 9 & 0 & 85 \\ 
  Cluster 23 & 2 & 18 & 2 & 25 & 22 & 69 \\ 
  Cluster 24 & 1 & 38 & 45 & 1 & 0 & 85 \\ 
  Cluster 25 & 2 & 57 & 16 & 1 & 0 & 76 \\ 
  Cluster 26 & 5 & 3 & 23 & 19 & 0 & 50 \\ 
  Cluster 27 & 9 & 0 & 30 & 5 & 20 & 64 \\ 
  Cluster 28 & 0 & 7 & 23 & 0 & 38 & 68 \\ 
  Cluster 29 & 27 & 2 & 10 & 1 & 11 & 51 \\ 
  Cluster 30 & 3 & 5 & 41 & 1 & 10 & 60 \\ \hline
  Total & 550 & 590 & 654 & 634 & 517 & 2945 \\ \hline
\end{tabular}
\end{table}

>>>>

In the \citet{tavazoie1999systematic} analysis, those authors performed
$k$-means clustering with $k = 30$; they found 23 of the clusters to be
uninterpretable, and they found 7 clusters to be meaningful. To compare our
clusters with the Tavazoie et al.\ clusters, we prepared a confusion matrix
comparing our clusters with the 7 interpretable Tavazoie clusters in
Table~\ref{table:confusion}.
%%(Appendix~\ref{app:confusion} contains a confusion matrix for all 30
%% Tavazoie clusters, including the non-interpretable ones.)
Entry $(i,j)$ of the confusion matrix gives the number of genes in
Tavazoie's Cluster~$i$ and our Cluster~$j$.

@@

In this paper, we proposed a new approach to estimate the number of clusters
to be used in $k$-means clustering. The intuition behind our proposed method
is to transform the unsupervised learning problem into a supervised learning
problem via a form of Gabriel cross validation.  We proved that our method is
self-consistent, and we analyzed its behavior in some special cases of
Gaussian mixture models.  Using simulations and real data examples, we
demonstrated that our method has good performance, competitive with existing
approaches. The simulations and empirical benchmarks demonstrate the
advantages of our method. In the yeast cell cycle application, our method
was able to identify meaningful gene groups with a small number of clusters.

There are many other clustering algorithms that get used in practice besides
$k$-means. We suspect that it should be possible to apply our method in the
context of a spectral clustering procedure, after transforming by the
eigenvectors of the Laplacian matrix. For other clustering schemes, including
versions of hierarchical clustering, we are less certain about the viability
of Gabriel cross-validation.  It is an open question as to whether Gabriel
cross-validation can be extended to other clustering methods, and whether such
extensions will perform well in practice.

For $k$-means clustering, Gabriel cross-validation is competitive with
other model selection methods, especially in the presence of high-dimensional,
heterogeneous, or heavy-tailed data.

%% Besides no strong modeling assumption is required, our proposed method is
%% robust against data set with variance heterogeneity, unequal number of
%% observations, non-Gaussian noise and high-dimension.  Such robustness is
%% important in practice because for any given data, it's hard to tell whether
%% its clusters have different number of observations, is the variance equal for
%% each cluster, or what underlying noise distribution it has. The weakness of
%% our proposed method is that its theory assumes only week correlation between
%% ``predictor'' columns and ``response'' columns. In practice, many data sets
%% don't exhibit high correlations between columns, where our proposed method can
%% be safely applied. In case the high correlation does exist, the
%% correlation-corrected version of Gabriel cross-validation method in section
%% $4.3$ can be used if we can assume common covariance structure. Other
%% procedure such as leave out most columns for clustering may also be used to
%% reduce its effect. However, the theory for such procedure has not been fully
%% developed and it could be the future research topic. 
%% 
%% 
%% Another situation where our proposed method cannot be directly used is when
%% the clusters are non-convex. In such situation, $k$-means itself doesn't work
%% well, for example two concentric circles share the same cluster center
%% \citep{hastie2009elements}. However, it's possible that our proposed Gabriel
%% cross-validation method can be used on the transformed data set. In the
%% concentric circles case where spectral clustering is appropriate, our proposed
%% method can be applied on the eigenvector subspace of the graph Laplacian
%% matrix inside the spectral clustering algorithm to find the optimal $k$. This
%% can also be the future research topic.

@@

\section{Clustering scree plot examples}

@@

%% \section{Review of consistency results for clustering}
%% 
%% Some proposed methods provide theoretical consistency result for choosing $k$.
%% Assume data is a mixture of $G$ $p$-dimension clusters with equal prior,
%% identically distributed with common covariance matrix and finite fourth
%% moments in each dimension, \citet{sugar2003finding} shows the Jump method will
%% pick the correct $k$ under the conditions that cluster centers are
%% sufficiently separated and an appropriate transformation is used.
%% Specifically, let $\Delta\sqrt{p}$ denotes the minimum Euclidean distance
%% between cluster centers, \citet{sugar2003finding} shows that the jump will be
%% maximized at $k=G$ given that $\Delta > 6$ and the existence of a positive $Y$
%% such that
%% \[
%%   \left(\frac{p\Delta^2W}{9G}\right)^{-Y}
%%     + \left(
%%         W \left[\frac{2^{2H^*(X)}}{K^2_{max}2\pi e}\right]
%%         - (\frac{\Delta}{6})^2(1-W)
%%       \right)^{-Y} < 2
%% \]
%% and $\left(\frac{p\Delta^2W}{9G}\right)^{-Y} <1/2$, where $H^*(X)$ is the
%% minimum entropy of the cluster membership over each dimension and $W =1-
%% \frac{6^4V_{\mathbf{X}}}{(\Delta^2-36)^2}$ with $V_{\mathbf{X}} = \var\left(
%%   1/p ||\mathbf{X}-\mu_j||^2_{\Gamma^{-1}} \mid \mathbf{X}
%% \hspace{0.05in}\text{in jth cluster}\right)$
%% 
%% \cite{tibshirani2005cluster} proved that the prediction strength method is
%% consistent under the setting that the observations in $p$-dimension are
%% uniformly distributed within one unit from their respective cluster centers,
%% with minimum distance between the $G$ cluster centers is four unit. Under such
%% well-separated setting, they show that 
%% \[ ps(G) = 1+ o_p(1), \hspace{0.5in} \underset{k \geq G+1}{\text{sup}} ps(k)\leq \frac{2}{3}+ o_p(1) \]
%% therefore $\hat{k}$ is consistent in estimating $G$.  When data does come from
%% mixture of Gaussian distribution, the model-based method of
%% \citet{fraley2002model} is consistent in estimating $k$ in low dimension.
%% 
%% Let $\Psi$ denotes any given base-clustering algorithm (e.g $k$-means), and
%% $\Psi(Z,k)$ the clustering obtained by applying $\Psi$ on data $Z$ with
%% parameter $k$. Also let $k_0$ be the optimal number of cluster and
%% $d\{\Psi(Z_1,k),\Psi(Z_2,k)\}$ be the measure of distance between clustering
%% $\Psi(Z_1,k)$ and $\Psi(Z_2,k)$ where $Z_1$ and $Z_2$ are two independent
%% samples from population. \citet{wang2010consistent} shows that their proposed
%% method will consistently select $k = k_0$, given such $k_0$ exists and
%% $d\{\Psi(Z_1,k),\Psi(Z_2,k)\}$ converges to zero at proper rate $r_k$.
%% However, $k_0$ is defined as the most stable number for $\Psi$ among all the
%% possible $k$, instead of the number of components in mixture model or
%% well-separated compact regions commonly seen in literature. Therefore, their
%% consistent result actually is convergent result, i.e. the selected $k$
%% converges to its limit $k_0$ given such limit exists.

@@

%% \section{Confusion matrix} \label{app:confusion}
%% 
%% \begin{table}
%%   \footnotesize
%% %\captionsetup{justification=centering}
%% \caption{\label{table:confusion-complete} Complete confusion matrix comparing
%% $5$ clusters to $30$ clusters }
%% \begin{tabular}{lccccc|c}
%%   \toprule
%%  & Cluster 1 & Cluster 2 & Cluster 3 & Cluster 4 & Cluster 5 & Total \\ 
%%   \midrule
%%   Cluster 1 & 0 & 0 & 1 & 161 & 2 & 164 \\ 
%%   Cluster 2 & 1 & 0 & 0 & 0 & 185 & 186 \\ 
%%   Cluster 3 & 0 & 0 & 91 & 11 & 2 & 104 \\ 
%%   Cluster 4 & 0 & 102 & 2 & 66 & 0 & 170 \\ 
%%   Cluster 5 & 152 & 0 & 0 & 0 & 0 & 152 \\ 
%%   Cluster 6 & 1 & 2 & 12 & 85 & 4 & 104 \\ 
%%   Cluster 7 & 1 & 10 & 83 & 7 & 0 & 101 \\ 
%%   Cluster 8 & 3 & 145 & 0 & 0 & 0 & 148 \\ 
%%   Cluster 9 & 142 & 0 & 0 & 0 & 5 & 147 \\ 
%%   Cluster 10 & 12 & 27 & 24 & 20 & 6 & 89 \\ 
%%   Cluster 11 & 0 & 5 & 81 & 8 & 0 & 94 \\ 
%%   Cluster 12 & 3 & 68 & 5 & 3 & 1 & 80 \\ 
%%   Cluster 13 & 0 & 1 & 12 & 81 & 5 & 99 \\ 
%%   Cluster 14 & 0 & 1 & 29 & 6 & 38 & 74 \\ 
%%   Cluster 15 & 60 & 53 & 2 & 0 & 0 & 115 \\ 
%%   Cluster 16 & 7 & 0 & 0 & 9 & 83 & 99 \\ 
%%   Cluster 17 & 9 & 0 & 72 & 1 & 1 & 83 \\ 
%%   Cluster 18 & 3 & 16 & 10 & 67 & 5 & 101 \\ 
%%   Cluster 19 & 62 & 2 & 0 & 1 & 8 & 73 \\ 
%%   Cluster 20 & 11 & 0 & 6 & 2 & 65 & 84 \\ 
%%   Cluster 21 & 1 & 13 & 6 & 44 & 6 & 70 \\ 
%%   Cluster 22 & 33 & 15 & 28 & 9 & 0 & 85 \\ 
%%   Cluster 23 & 2 & 18 & 2 & 25 & 22 & 69 \\ 
%%   Cluster 24 & 1 & 38 & 45 & 1 & 0 & 85 \\ 
%%   Cluster 25 & 2 & 57 & 16 & 1 & 0 & 76 \\ 
%%   Cluster 26 & 5 & 3 & 23 & 19 & 0 & 50 \\ 
%%   Cluster 27 & 9 & 0 & 30 & 5 & 20 & 64 \\ 
%%   Cluster 28 & 0 & 7 & 23 & 0 & 38 & 68 \\ 
%%   Cluster 29 & 27 & 2 & 10 & 1 & 11 & 51 \\ 
%%   Cluster 30 & 3 & 5 & 41 & 1 & 10 & 60 \\ \midrule
%%   Total & 550 & 590 & 654 & 634 & 517 & 2945 \\ \bottomrule
%% \end{tabular}
%% \end{table}
